Rheology of Cubic Blue Phases

### 2012-09-04 papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1776 40876441-3496-0410-a613-a2fca35b2ae2

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,pre,reprint,superscriptaddress]{revtex4}
%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}

@@

Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by
$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
order. More generally,
the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
characterizes the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2-\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.

A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
which are given by
\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
\end{eqnarray}
The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{eqn2}
\end{equation}
The first term on the left hand side of Eq.\ref{eqn2} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
\mathrm{Tr}({\mathbf Q W}),
\label{eqn3}
\end{eqnarray}
where $\mathrm{Tr}$ denotes the tensorial trace, while 
${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by 
setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{eqn4}
\end{equation}
The rotational diffusion constant $\Gamma$ in Eq.~\ref{eqn2} is proportional
to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
\cite{deGennes}.

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
= \partial_\beta \Pi_{\alpha \beta}
+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
\label{eqn6}
\end{eqnarray}
The final term in Eq.~\ref{eqn6} arises from the Chapman-Enskog expansion
of the LB equations \cite{Denniston:2001}.
At low flow rates the fluid can be considered as incompressible, so that the
last term on the right hand side of Eq.~\ref{eqn6} remains small.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
scalar pressure which, in a system at rest, is constant to a very good
approximation.

The system of coupled partial differential equations Eqs.~\ref{eqn2}
and~\ref{eqn6} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
with predictor-corrector scheme for the continuity equation and
Eq.~\ref{eqn6}, and a finite difference scheme for the equation of motion of
the tensor order parameter Eq.\ref{eqn2}. More details on the algorithm can
be found in~\cite{Denniston:2001, Denniston:2004}.

@@

@@

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions impose flow in such a way that network moves 
to the right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure*}

@@

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure*}

@@

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

@@

\begin{figure}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The configuration is a helical wave that travels along the direction of vorticity. It is translationally invariant in gradient and flow direction.}
\label{bp2-high}
\end{figure}
\clearpage

@@

In BPI the disclination lines are arranged in a body-centered-cubic array.
Contrary to BPII, they are well separated and do not intersect. 
This difference in topology leads to some stiking differences in the flow behaviour of both blue phases 
as we will see in the following.

We can identify three different regimes in the range of applied shear rates:
At high shear rates a breakup of the network structure occurs, similar to that seen in BPII. 
At intermediate shear rates BPI features seemingly chaotic but periodically recurring conformations.
At low shear rates the network undergoes first an affine transformation, but becomes unstable
at larger strains and adopts an amorphous configuration with well separated disclination lines.

@@

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. The network is displaced during every cycle, but the magnitude and direction of the offset are seemingly chaotic.}
\label{bp1-med}
\end{figure*}

The apparent viscosity of BPI in simple shear flow, depicted in Fig. \ref{bp1-rheo}, is larger for The degree of shear thinning, also visible in the inset in terms of average shear stress over
flow rate

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp1-rheo}
\end{figure}

Runs at smaller system sizes suggest that the individual conformation 
can depend sensitively on the system size. Fig. \ref{bp1-2uc4uc}

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
\caption{}
\label{bp1-2uc4uc}
\end{figure}

\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N)
\end{equation}

>>>>

%% ****** Start of file apstemplate.tex ****** %
%%
%%
%%   This file is part of the APS files in the REVTeX 4 distribution.
%%   Version 4.1r of REVTeX, August 2010
%%
%%
%%   Copyright (c) 2001, 2009, 2010 The American Physical Society.
%%
%%   See the REVTeX 4 README file for restrictions and more information.
%%
%
% This is a template for producing manuscripts for use with REVTEX 4.0
% Copy this file to another name and then work on that file.
% That way, you always have this original template file to use.
%
% Group addresses by affiliation; use superscriptaddress for long
% author lists, or if there are many overlapping affiliations.
% For Phys. Rev. appearance, change preprint to twocolumn.
% Choose pra, prb, prc, prd, pre, prl, prstab, prstper, or rmp for journal
%  Add 'draft' option to mark overfull boxes with black boxes
%  Add 'showpacs' option to make PACS codes appear
%  Add 'showkeys' option to make keywords appear
\documentclass[aps,pre,reprint,superscriptaddress, twocolumn]{revtex4}
%\documentclass[aps,pre,preprint,superscriptaddress]{revtex4}
%\documentclass[aps,prl,reprint,groupedaddress]{revtex4}

@@

Our approach is based on the well-established Beris-Edwards model for hydrodynamics of
cholesteric liquid crystals \cite{Beris:1994}, which describes the ordered state 
in terms of a traceless, symmetric tensor order parameter ${\mathbf Q}({\mathbf r})$. 
In the uniaxial approximation, the order parameter is given by
$Q_{\alpha \beta}= q_s ( n_\alpha n_\beta - \frac{1}{3}\; \delta_{\alpha\beta})$
with ${\mathbf n}$ the director field and $q_s$ the amplitude of nematic
order. More generally,
the largest eigenvalue of ${\mathbf Q}$, $0\le q_s\le\frac{2}{3}$
characterizes the local degree of orientational order.
The thermodynamic properties of the liquid crystal are determined by a free energy
${\cal F}$, whose density $f$ consists of a bulk contribution $f_b$ and a gradient part $f_g$, as follows,
\begin{eqnarray}
f_b&=&\frac{A_0}{2}\left(1-\frac{\gamma}{3}\right) Q_{\alpha \beta}^2\nonumber\\
&-&\frac{A_0 \gamma}{3}Q_{\alpha \beta} Q_{\beta \gamma} Q_{\gamma \alpha}+\frac{A_0 \gamma}{4}(Q_{\alpha \beta}^2)^2,\nonumber\\
f_g&=&\frac{K}{2}(\varepsilon_{\alpha\gamma\delta} \partial_\gamma Q_{\delta\beta}+2 q_0 Q_{\alpha \beta})^2+\frac{K}{2}(\partial_\beta Q_{\alpha \beta})^2.\label{FE}
\end{eqnarray}
The first term contains the bulk-free energy constant $A_0$ and the inverse temperature $\gamma$ which controls the magnitude of order.
The second part quantifies the cost of elastic distortions, which are proportional to the elastic constant $K$;
we work for simplicity in the one-elastic constant approximation. The wavevector $q_0=2\pi/p_0$, where $p_0$ is the cholesteric pitch.
The actual periodicity of the BP structure, $p$, does not need to be equal to $p_0$.
The ``redshift'' $r=p/p_0$ is adjusted during the simulation by following a simple procedure similar to~\cite{Alexander:2006}.

A particular thermodynamic state is specified by two dimensionless quantities: the effective temperature $\tau$ and chirality $\kappa$,
which are given by
\begin{eqnarray}
\tau&=&\frac{27(1-\gamma/3)}{\gamma}\nonumber\\
\kappa&=&\sqrt{\frac{108 K q_0^2}{A_0 \gamma}}\nonumber.
\end{eqnarray}
The dynamical evolution of the order parameter is given by the equation 
\begin{equation}
\left(\partial_t+ u_\alpha \partial_\alpha \right){\mathbf Q} - {\mathbf S}({\mathbf W},{\mathbf Q}) = \Gamma {\mathbf H}.
\label{op-eom}
\end{equation}
The first term on the left hand side of Eq.\ref{op-eom} is a material derivative, which describes the rate of change of a quantity moving along with the flow.
The second term accounts for the rate of change due to local velocity gradients $W_{\alpha \beta}=\partial_\beta u_\alpha$,
\begin{eqnarray}
{\mathbf S}({\mathbf W}, {\mathbf Q}) &=& (\xi {\mathbf A} + {\boldsymbol \Omega})({\mathbf Q}+\frac{\mathbf I}{3})\nonumber\\
& &\hspace*{-1.5cm}+ ({\mathbf Q}+\frac{\mathbf I}{3})(\xi {\mathbf A}  - {\boldsymbol \Omega})-2 \xi ({\mathbf Q}+\frac{\mathbf I}{3})
\mathrm{Tr}({\mathbf Q W}),
\label{sw}
\end{eqnarray}
where $\mathrm{Tr}$ denotes the tensorial trace, while 
${\mathbf A}=({\mathbf W}+{\mathbf W}^T)/2$ and
${\boldsymbol \Omega}=({\mathbf W}-{\mathbf W}^T)/2$ are the symmetric and antisymmetric part of the velocity gradient, respectively. $\xi$ 
is a constant depending on the molecular details of the liquid crystal.
Flow alignment occurs if $\xi \cos{2\theta}=(3q_s)/(2+q_s)$ has a real solution, where $\theta$ is the Leslie-angle: we select this case by 
setting $\xi=0.7$ in our simulations.
${\mathbf H}$ is the molecular field, which is a functional derivative of $\cal F$ that preserves the tracelessness of $\mathbf Q$:
\begin{equation}
{\bf H}=-\frac{\delta {\cal F}}{\delta {\bf Q}}+\frac{\bf I}{3}\,
\mathrm{Tr} \left(\frac{\delta {\cal F}}{\delta {\bf Q}}\right).
\label{molfield}
\end{equation}
The rotational diffusion constant $\Gamma$ in Eq.~\ref{op-eom} is proportional
to the inverse of the rotational viscosity $\gamma_1=2 q_s^2/\Gamma$
\cite{deGennes}.

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
&=& \partial_\beta \Pi_{\alpha \beta}+ \eta\, \partial_\beta \{ \partial_\alpha u_\beta \nonumber\\
& &\hspace*{-1.5cm} +\; \partial_\beta u_\alpha
+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
\label{NSE}
\end{eqnarray}
The final term in Eq.~\ref{NSE} arises from the Chapman-Enskog expansion
of the LB equations \cite{Denniston:2001}.
At low flow rates the fluid can be considered as incompressible, so that the
last term on the right hand side of Eq.~\ref{NSE} remains small.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in LBU.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)\nonumber\\
&-&\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta} + Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta} \nonumber\\
&+&2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}
- \partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
\label{Pi}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{Pi} is reduced to the
scalar pressure which, in a system at rest, is constant to a very good
approximation.

The spectrum $X_\omega$ of the deviatoric stress $\Pi_{xy}(t)$ was determined via discrete Fourier transformation accoring to
\begin{equation}
X_\omega=\frac{1}{\sqrt{N}}\sum_{n=0}^{N-1} \left[\Pi_{xy}(t_n) - \bar{\Pi}_{xy}\right] \exp(2\pi i \, \omega \,n\,/\,N),
\label{spectrum}
\end{equation}
wheras we subtracted the average stress per cycle $\bar{\Pi}_{xy}$ from the signal.
$N$ was the number of points in the time series, typically one for every 100 LB time steps.

The system of coupled partial differential equations Eqs.~\ref{op-eom}
and~\ref{NSE} is solved by means of a
hybrid scheme~\cite{Marenduzzo:2007}. This uses a lattice Boltzmann algorithm
with predictor-corrector scheme for the continuity equation and
Eq.~\ref{NSE}, and a finite difference scheme for the equation of motion of
the tensor order parameter Eq.\ref{op-eom}. More details on the algorithm can
be found in~\cite{Denniston:2001, Denniston:2004}.

@@

\begin{figure}[ht!]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure}

@@

\begin{figure}[ht]
\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions impose flow in such a way that network moves 
to the right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

@@

@@

\begin{figure*}[ht]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

@@

\begin{figure*}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The configuration is a helical wave that travels along the direction of vorticity. It is translationally invariant in gradient and flow direction.}
\label{bp2-high}
\end{figure*}

@@

In BPI the disclination lines are arranged in a body-centered-cubic array.
Contrary to BPII, they are well separated and do not intersect. 
There is strong evidence that these topological differences result in 
a very different rheological response of BPI. In fact BPI exhibits some 
peculiarities that are absent in BPII.

Just as before in a kind of overview we present rheological key quantities first and 
will address the detail aspects further below. 
Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
the apparent viscosity varies by almost a factor 8 during one cycle.
The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
follow a simple power law.

The inset of shows the flow curve, i.e. shear stress over shear rate.
The data points are averages over one cycle, whereas the error bars indicate the maximum and
minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
At high shear rates the network breaks up similar to BPII. The director field in this state
is effectively two-dimensional and translationally invariant along the neutral direction.
At intermediate shear rates the BPI network flows with periodically recurring conformations.
These complex flow-induced conformations are accompanied by large fluctuations that increase 
in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
of the liquid crystal during one cycle is actually negative. 
Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle). The error bars give the minimum and maximum shear stress during one cycle.}
\label{bp1-rheo}
\end{figure}

The Fourier analysis of the time-dependent deviatoric stress $\Pi_{xy}$ 
offers another perspective on this intricate flow behaviour at high and intermediate shear rates. 
The spectrum $X_\omega$ according to Eq. \ref{spectrum} is shown in Fig. \ref{bp1-spectrum}.
We observe a strong contribution of the first harmonic, which is directly related to the 
recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ 
via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
As the shear rates we applied always differ by a factor 2 the Fourier coefficients of the 
individual runs with different shear rate coincide in such a way that the
the fundamental mode is the same frequency as the first overtone of the next lower shear rate
or the third overtone of the next-to-next lower shear rate 
(viz. peak positions at $\gd=7.81, 3.91$ and $1.95\e{-5}$).
Moreover, the magnitude of higher Fourier modes decreases monotonously.   
However, the inset shows that this characteristic frequency relation is violated 
for shear rates above $\gd=7.81\e{-5}$.
Although the velocity of the primary flow component in x-direcition, listed in Tab. \ref{tab1},
 is very well in accordance with the imposed shear rate, the fundamental mode appears 
now at the same frequency.
The contribution of higher Fourier modes increases and even exceeds that of
the fundamental mode, which gives the shear stress generally a more jittering appearance.
At the highest imposed shear rate where the breakup occurs the Fourier spectrum 
becomes simple again and features a dominating fundamental mode at the 
expected frequency just as described above for $\gd\le7.81\e{-5}$.
The breakup occurs at exactly the same flow rate as those of BPII, 
suggesting that this can be traced back to the rotational dynamics 
of the order parameter.

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate. The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed. Instead the spectral contribution of higher modes grows until finally the breakup takes place.}
\label{bp1-spectrum}
\end{figure}

The rheological response of BPI at low shear rate is strikingly 
different from that of BPII.
We want to explain these differences on the basis of a typical 
appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
Three different stages can be distinguished. 
With the onset of shearing the disclination lines 
in BPI get more and more squeezed together which is incommensurable with 
the defect topology of this blue phase. 
Consequently, the network adopts a flow-induced conformation which consists 
of intertwined helices that stretch during the affine transformation.
This helical conformation that emerges already at strains $\gamma\simeq1$ 
is shown in Fig. \ref{bp1-low} (top row).
At a particular point in the cycle a non-helical configuration is
observed (bottom left) before the network quickly returns
to its helical state.
There is no perceivable movement of the network in vorticity 
direction at any stage of the cycle.

However, this mode of flow proves unstable as after a few cycles 
distortions appear which lead to further destabilisation.
Finally, the network

\begin{figure*}[ht]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$).}
\label{bp1-low}
\end{figure*}

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
\caption{}
\label{bp1-fe-yield}
\end{figure}

@@

\begin{figure*}[ht]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. The network is displaced during every cycle, but the magnitude and direction of the offset are seemingly chaotic.}
\label{bp1-med}
\end{figure*}

\begin{figure*}[ht]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero. The colour code gives the magnitude and sign of the z-component. The vertical and horizontal direction are the gradient and vorticity direction, respectivelty. Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
\label{bp1-velo}
\end{figure*}

Runs at smaller system sizes suggest that the individual conformation 
can depend sensitively on the system size. Fig. \ref{bp1-low-fe}

\begin{figure}[ht]
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\caption{}
\label{bp0-2uc4uc-fe}
\end{figure}

\begin{figure}[ht]

\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{bp1-spec}
\end{figure}

The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.
